\large{\bf Looking forwards and backwards through locally regulated populations }

### 2022-04-22 forwards-and-backwards.tex
.

Concretely, then: for each label $a$ in
$\labelspace = \bigcup_{k \ge 1} \IN_{\ge 1}^k$
let $\Pi_a$ be an independent Poisson process on $[0, \infty)^2 \times \IR_d \times \{0,1\}$.
The mean measure is a product of Lesbegue measure on the first two coordinates,
$q(0, \cdot)$ on $\IR^d$, and $(\delta_0 + \delta_1)/2$ on $\{0, 1\}$.
It will also be convenient
to suppose that for each label $a$ we have an enumeration of the points in $\Pi_a$,
so we may refer to ``the $j^\text{th}$ point in $\Pi_a$'',
although the precise order of this enumeration is irrelevant.
If $(t, v, y, \kappa)$ is the $j^\text{th}$ point in $\Pi_a$,
then $t$ will determine possible birth times,
$v$ will determine levels of the offspring,
$y$ will give the spatial displacement of the offspring relative to the parent,
$\kappa$ will be used to determine whether parent or offspring are assigned the new level,
and the new label produced, if any will be $aj$,
i.e., the label $a$ with $j$ appended (so, if $a \in \IN_{\ge 0}^k$ then $aj \in \IN_{\ge 0}^{k+1}$).
Each label $a$ has a birth time $\tau_a$,
when it is first assigned,
and a death time $\sigma_a$, when its level first hits $\lambda$.
For any $\tau_a \le t \le \sigma_a$ we denote by $X_a(t)$ and $U_a(t)$ the spatial location and level
of the individual carrying label $a$ at time $t$.
Now, since we have defined labels so that the level does not jump,
$U_a$ satistifies~\eqref{eqn:dot_u} for $\tau_a \le t \le \sigma_a$, i.e.,
\begin{equation*}
    \begin{split}
& U_a(t)
    =
    U_a(\tau_a) \\
&\qquad {}   
    + \int_{\tau_a}^{t}
    \left(
        \frac{\theta}{\lambda} \gamma(X_a(s),\eta(s))
        \int_{\IR^d} r(z,\eta(s)) q(X_a(s),dz) U_a(s)^2
        -
        b(X_a(s),\eta(s)) U_a(s)
    \right)
    ds ,
\end{split}
\end{equation*}
and, of course, $\sigma_a = \inf\{t \ge \tau_a : U_a(t) > \lambda\}$.

Potential reproduction events occur at times $\tau_a + s$
for each point $(s, v, y, \kappa) \in \Pi_a$.
We say `potential' since if the resulting level is greater than $\lambda$,
the event does not happen.
If this is the $j^\text{th}$ point in $\Pi_a$,
the new label is $aj$, the (potential) birth time is $\tau_{aj} = \tau_a + s$.
For convenience define
$$
    \ell(x, y, \eta, v)
    =
    \frac{
        v
    }{
        2 \lambda^{-1} \theta \gamma(x, \eta) r(x + y, \eta)
    } ,
$$
and using this, the (potential) new level is
\begin{equation*}
    U_{aj}(\tau_{aj})
    =
    U_a(\tau_{aj})
    +
    \ell(X_a(\tau_{aj}-), y, \eta(\tau_{aj}-), v) .
\end{equation*}
If $U_{aj}(\tau_{aj}) < \lambda$,
the new individual labeled $aj$ is produced,
and $\kappa$ determines which label is associated with the new location,
so
\begin{align*}
    X_{aj}(\tau_a + s)
    &=
    X_a(\tau_a+s-) + (1 - \kappa) y .
\end{align*}
On the other hand
if $U_{aj}(\tau_a+s) \ge \lambda$, then $X_a$ is unchanged and $X_{aj}$ is undefined,
so
\begin{align*}
    X_a(\tau_a+s)
    &=
    X_a(\tau_a+s-) + \kappa y \mathbbm{1}_{U_{aj}(\tau_{aj}) < \lambda} .
\end{align*}
Recall that the parental \emph{individual} always retains their spatial location,
so that $\kappa = 0$ corresponds to the parent being assigned a new level,
and our line of descent switching to the offspring.

@@

\comment{Do we want subscript $a$s here, on e.g., $\zeta$?
I think not, because this is all given a particular $a$?}

First, for each $1 \le k < n$
let $(s_k, v_k, y_k, \kappa_k)$ be the $a_{k+1}^\text{th}$ point in $\Pi_{(a)_k}$
(this is the sequence of jumps taken by the line of descent).
Note that $s_k = \tau_{(a)_k}$, and also set $s_0 = 0$.
Let $\Gamma_a$ by the point process ``experienced by the line of descent leading to $a$'',
which is easier to define if we treat point processes as sets for a moment,
defining
\begin{align*}
    \Gamma_a
    &=
    \bigcup_{k=1}^n 
    \bigg\{
        (s, v, y, \kappa) \in \Pi_{(a)_k} \; : \;
        \tau_{(a)_{k-1}} \le s < \tau_{(a)_{k-1}} 
    \\& \qquad \qquad 
        \text{ and }
        U_{(a)_k}(\tau_{(a)_{k-1}} + s) + \ell( X_{(a)_k}(\tau_{(a)_{k-1}} + s-), y, \eta(\tau_{(a)_{k-1}} + s-), v) < \lambda
    \bigg\}
\end{align*}
\comment{TODO include the switch points}

which is for $A \subset [0,\infty)^2 \times \IR^d \times \{0,1\}$
\comment{This is wrong, as it includes non-viable individuals.}
\begin{align*}
\Gamma_a(A)
    &=
    \sum_{k=1}^n
    \Pi_{(a)_k}\left(A \cap [s_{k-1}, s_k) \times [0, \infty) \times \IR^d \times \{0,1\}\right) 
    \\ &\qquad {}
    + \sum_{k=1}^{n-1}
      \delta_{(s_k, v_k, y_k, 1 - \kappa_k)}(A) .
\end{align*}
Note that we have included the points at which the line of descent's level jumps,
flipping $\kappa$ at these points,
and as a result $\Gamma_a$ has the same distribution as $\Pi_a$
restricted to $[0, \sigma_a)$.
Then for $0 \le t \le \sigma_a$,
\begin{align} \label{eqn:X_line_of_descent}
    X_a(t)
    &=
    X_{a_1}(0)
    +
    \int_{
        [0, t) \times [0, \infty) \times \IR^d \times \{0, 1\}
    }
    \kappa y
    \mathbbm{1}_{
        U_{(a)_{\zeta(s)}}(s) + \ell( X_{(a)_{\zeta(s)}}(s-), y, \eta(s-), v) < \lambda
    }
    \Gamma_{a}(ds, dv, dy, d\kappa) 
\end{align}
and
\begin{align} \label{eqn:U_line_of_descent}
    U_a(t)
    &=
    U_{a_1}(0)
    +
    \sum_{k=1}^{\zeta_a(t) - 1}
    \frac{
        v_k
    }{
        2 \lambda^{-1} \theta \gamma(X_{(a)_k}(s_k-), \eta(s_k-))
        r(X_{(a)_k}(s_k-) + y_k, \eta(s_k-))
    }
    \\ & \qquad {}
    +
    \int_0^t
    \bigg(
        \frac{\theta}{\lambda} \gamma(X_a(s),\eta(s))
        \int_{\IR^d} r(z,\eta(s)) q(X_a(s),dz) U_a(s)^2
    \\ & \qquad\qquad  \qquad {}
        -
        b(X_a(s),\eta(s)) U_a(s)
    \bigg)
    dt .
\end{align}

@@

The previous section constructs the lookdown process
in a way that makes individual lines of descent converge pointwise as $\lambda \to \infty$,
using the same underlying Poisson processes $\{\Pi_a\}_{a \in \labelspace}$.
To see this, first note that if the Poisson processes are fixed
then the set of events to which a given label $a \in \labelspace$ is associated
is also fixed -- this is the sequence $(s_k, v_k, y_k, \kappa_k)$ associated with a given label $a$.
Therefore, the processes $\Gamma_a$ are also unchanged.
First, we clearly need that the spatial measures $\eta$ converge.
\comment{(hm, no notation for $\eta$ depending on $\lambda$ here; need to introduce this here?)}
Now, consider equations~\eqref{eqn:X_line_of_descent} and~\eqref{eqn:U_line_of_descent}.

Equation~\eqref{eqn:X_line_of_descent} simply says that $X_a$ follows a random walk

>>>>

Concretely, then: for each label $a$ in
$\labelspace = \bigcup_{k \ge 1} \IN_{\ge 1}^k$
let $\Pi_a$ be an independent Poisson process on $[0, \infty)^2 \times \IR_d \times \{0,1\}$.
The mean measure is a product of Lesbegue measure on the first two coordinates,
$q(0, \cdot)$ on $\IR^d$, and $(\delta_0 + \delta_1)/2$ on $\{0, 1\}$.
It will also be convenient
to suppose that for each label $a$ we have an enumeration of the points in $\Pi_a$,
so we may refer to ``the $j^\text{th}$ point in $\Pi_a$'',
although the precise order of this enumeration is irrelevant.
If $(t, v, y, \kappa)$ is the $j^\text{th}$ point in $\Pi_a$,
then $t$ will determine possible birth times,
$v$ will determine levels of the offspring,
$y$ will give the spatial displacement of the offspring relative to the parent,
$\kappa$ will be used to determine whether parent or offspring are assigned the new level,
and the new label produced, if any will be $aj$,
i.e., the label $a$ with $j$ appended (so, if $a \in \IN_{\ge 0}^k$ then $aj \in \IN_{\ge 0}^{k+1}$).
Each label $a$ has a birth time $\tau_a$,
when it is first assigned,
and a death time $\sigma_a$, when its level first hits $\lambda$.
For any $\tau_a \le t \le \sigma_a$ we denote by $X_a(t)$ and $U_a(t)$ the spatial location and level
of the individual carrying label $a$ at time $t$.
Now, since we have defined labels so that the level does not jump,
$U_a$ satistifies~\eqref{eqn:dot_u} for $\tau_a \le t \le \sigma_a$, i.e.,
\begin{equation} \label{eqn:U_line_of_descent}
    \begin{split}
& U_a(t)
    =
    U_a(\tau_a) \\
&\qquad {}   
    + \int_{\tau_a}^{t}
    \left(
        \frac{\theta}{\lambda} \gamma(X_a(s),\eta(s))
        \int_{\IR^d} r(z,\eta(s)) q(X_a(s),dz) U_a(s)^2
        -
        b(X_a(s),\eta(s)) U_a(s)
    \right)
    ds ,
\end{split}
\end{equation}
and, of course, $\sigma_a = \inf\{t \ge \tau_a : U_a(t) > \lambda\}$.

Potential reproduction events occur at times $\tau_a + s$
for each point $(s, v, y, \kappa) \in \Pi_a$.
We say `potential' since if the resulting level is greater than $\lambda$,
the event does not happen.
If this is the $j^\text{th}$ point in $\Pi_a$,
the new label is $aj$, the (potential) birth time is $\tau_{aj} = \tau_a + s$.
For convenience define
$$
    \ell(x, y, \eta, v)
    =
    \frac{
        v
    }{
        2 \lambda^{-1} \theta \gamma(x, \eta) r(x + y, \eta)
    } ,
$$
and using this, the (potential) new level is
\begin{equation*}
    U_{aj}(\tau_{aj})
    =
    U_a(\tau_{aj})
    +
    \ell(X_a(\tau_{aj}-), y, \eta(\tau_{aj}-), v) .
\end{equation*}
If $U_{aj}(\tau_{aj}) < \lambda$,
the new individual labeled $aj$ is produced,
and $\kappa$ determines which label is associated with the new location,
so
\begin{align*}
    X_{aj}(\tau_a + s)
    &=
    X_a(\tau_a+s-) + (1 - \kappa) y .
\end{align*}
On the other hand
if $U_{aj}(\tau_a+s) \ge \lambda$, then $X_a$ is unchanged and $X_{aj}$ is undefined,
so
\begin{align} \label{eqn:X_line_of_descent}
    X_a(\tau_a+s)
    &=
    X_a(\tau_a+s-) + \kappa y \mathbbm{1}_{U_{aj}(\tau_{aj}) < \lambda} .
\end{align}
Recall that the parental \emph{individual} always retains their spatial location,
so that $\kappa = 0$ corresponds to the parent being assigned a new level,
and our line of descent switching to the offspring.

We can now express our population processes using these levels:
\begin{align*}
    \eta_t = \sum_{a : \tau_a > t \;\&\; U_a(t) < \lambda} \delta_{X_a(t)}
    \qquad \text{and} \qquad
    \lp = \sum_{a : \tau_a > t \;\&\; U_a(t) < \lambda} \delta_{(X_a(t), U_a(t))} .
\end{align*}

\comment{
    The following is quite annoying to get right, but can fixed up.
    However, I don't think we need it.}

@@

First, for each $1 \le k < n$
let $(s_k, v_k, y_k, \kappa_k)$ be the $a_{k+1}^\text{th}$ point in $\Pi_{(a)_k}$
(this is the sequence of jumps taken by the line of descent).
Note that $s_k = \tau_{(a)_k}$, and also set $s_0 = 0$.
Let $\Gamma_a$ by the point process ``experienced by the line of descent leading to $a$'',
which is for $A \subset [0,\infty)^2 \times \IR^d \times \{0,1\}$
\comment{This is wrong, because $s$ is relative to birth time.}
\begin{align*}
\Gamma_a(A)
    &=
    \sum_{k=1}^n
    \Pi_{(a)_k}\left(A \cap [s_{k-1}, s_k) \times [0, \infty) \times \IR^d \times \{0,1\}\right) 
    \\ &\qquad {}
    + \sum_{k=1}^{n-1}
      \delta_{(s_k, v_k, y_k, 1 - \kappa_k)}(A) .
\end{align*}
Note that we have included the points at which the line of descent's level jumps,
flipping $\kappa$ at these points,
and as a result $\Gamma_a$ has the same distribution as $\Pi_a$
restricted to $[0, \sigma_a)$.
Then for $0 \le t \le \sigma_a$,
\begin{align*}
    X_a(t)
    &=
    X_{a_1}(0)
    +
    \int_{
        [0, t) \times [0, \infty) \times \IR^d \times \{0, 1\}
    }
    \kappa y
    \mathbbm{1}_{
        U_{(a)_{\zeta(s)}}(s) + \ell( X_{(a)_{\zeta(s)}}(s-), y, \eta(s-), v) < \lambda
    }
    \Gamma_{a}(ds, dv, dy, d\kappa) 
\end{align*}
and
\begin{align*}
    U_a(t)
    &=
    U_{a_1}(0)
    +
    \sum_{k=1}^{\zeta_a(t) - 1}
    \frac{
        v_k
    }{
        2 \lambda^{-1} \theta \gamma(X_{(a)_k}(s_k-), \eta(s_k-))
        r(X_{(a)_k}(s_k-) + y_k, \eta(s_k-))
    }
    \\ & \qquad {}
    +
    \int_0^t
    \bigg(
        \frac{\theta}{\lambda} \gamma(X_a(s),\eta(s))
        \int_{\IR^d} r(z,\eta(s)) q(X_a(s),dz) U_a(s)^2
    \\ & \qquad\qquad  \qquad {}
        -
        b(X_a(s),\eta(s)) U_a(s)
    \bigg)
    dt .
\end{align*}

@@

The previous section constructs the lookdown process
in a way that makes individual lines of descent converge pointwise as $\lambda \to \infty$,
using the same underlying Poisson processes $\{\Pi_a\}_{a \in \labelspace}$.
To see this, first note that if the Poisson processes are fixed
then the set of events to which a given label $a \in \labelspace$ is associated
is also fixed -- this is the sequence $(s_k, v_k, y_k, \kappa_k)$ associated with a given label $a$.
Therefore, the processes $\Gamma_a$ are also unchanged.
First, we clearly need that the spatial measures $\eta$ converge.
\comment{(hm, no notation for $\eta$ depending on $\lambda$ here; need to introduce this here?)}

Supposing that they do,
consider how a line of descent $(X_a(t), U_a(t))$ evolves.
The line of descent throws off a new line of descent at a higher level
when there is a point $(s, v, y, \kappa)$ in $\Pi_a$ with 
\begin{equation} \label{eqn:line_points}
v < 2 (\lambda - U_a(t)) \lambda^{-1} \theta \gamma(X_a(t), \eta) r(X_a(t)+y, \eta) .
\end{equation}
Since the mean measure of the $v$ coordinate is Lebesgue,
$\theta/\lambda \to \delta$,
and $q(x, dy) \to \delta_x(dy)$,
this corresponds in the limit to new lines of descent being thrown off at all times,
i.e., according to a Poisson process with rate
$$
\delta \gamma(X_a(t), \eta) r(X_a(t), \eta) 
$$
and uniform intensity on higher levels.
Again, since in the limit the dispersal distance is zero,
new levels are produced at the same spatial location.
However, this affects the location of the line of descent:
for each new line of descent, there is a $1/2$ chance
that the line of descent switches location (i.e., moves to $X_a(t) + y$).
Taking $g$ to be a suitable test function on $\IR^d$, 
and rewriting~\eqref{eqn:line_points},
the generator of this motion is, when the level is $u$,
\begin{align*}
    g(x)
    &\mapsto
    \left(1 - \frac{u}{\lambda}\right) \gamma(x, \eta)
    \theta \int_{\IR^d} r(x+y, \eta) (g(x+y) - g(x)) q(x, dy) \\
    &=
    \left(1 - \frac{u}{\lambda}\right) \gamma(x, \eta)
    \bigg\{
        \theta \int_{\IR^d} (r(x+y, \eta) g(x+y) - r(x, \eta) g(x)) q(x, dy) \\
        &\qquad \qquad {}
        -
        \theta \int_{\IR^d} (r(x+y, \eta) - r(x, \eta) ) g(x) q(x, dy) 
    \bigg\} \\
    &\to
    \gamma(x, \eta)
    \left(
        \DG(rg)(x) - g(x) \DG(r)(x)
    \right) .
\end{align*}
Notice that the factors of 2 have canceled,
and that the result is independent of $u$.
